---
title: "Roquin1 hyperinflammation"
output: 
  html_document:
    toc: yes
  html_notebook:
    theme: united
    toc: yes
---

This is a computational flow analysis accompanying the paper by Tavernier, 
Athanasopoulos, Verloo et. al.  
  
It contains a FlowSOM analysis of 9 fcs files, as exported from a BD FACSymphony,
and accompanied by a manual gating analysis of the files in FlowJo.

# Project setup

## Libraries and helper functions

We make use of several libraries for this analysis.  
  
If not yet installed, they can be installed using the following code:

```{r}
# install.packages("BiocManager")
# BiocManager::install("flowCore")
# BiocManager::install("flowDensity")
# BiocManager::install("flowWorkspace")
# 
# install.packages("ggplot2")
# install.packages("RColorBrewer")
# 
# install.packages("devtools")
# devtools::install_github("saeyslab/FlowSOM")
```

Once installed, they must be loaded:

```{r}
suppressPackageStartupMessages(library(flowCore))
suppressPackageStartupMessages(library(flowDensity))
suppressPackageStartupMessages(library(flowWorkspace))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(FlowSOM))
```

We also used some helper functions which are not yet included in these libraries:

```{r}
source("helperFunctions.R")
```

## FCS file location

All fcs files and corresponding workspaces are assumed to be placed in a folder 
called "data". We identify the correct fcs files for the sample ids of interest.
```{r}
fcs_dir <- "data"

sample_ids <- c("HC249", "HC255", "HC256", "HC257",
                "HC258", "HC259", "HC262", "PID630")
controls <- grep("HC", sample_ids, value = TRUE)

fcs_files <- sapply(sample_ids, 
                    function(x){
                      grep(paste0(x, ".*fcs"), 
                           list.files(fcs_dir),
                           value = TRUE)
                    })
fcs_files
```

## Color definitions

We define specific colors for each sample (blue for HC, red for patient) and
for each cell type of interest.

```{r}
sample_colors <- c(RColorBrewer::brewer.pal(9,"Blues")[-c(1,2)], 
                "#d7301f")
names(sample_colors) <- sample_ids

celltype_colors <- c("Unknown" = "#FFFFFF",
                     "Basophils" = "#5854e2",
                     "cDCs" = "#e2da54",
                     "pDCs" = "#a90688",
                     "Monocytes" = "#f37611",
                     "Tregs" = "#ef82b8",
                     "Na CD4+ T cells" = "#97b867",
                     "Eff CD4+ T cells" = "#e2cb5f",
                     "MAIT cells" = "#632d0f",
                     "Na CD8+ T cells" = "#7c418f",
                     "gd T cells" = "#2b5c0d",
                     "DN T cells" = "#4fb214",
                     "Eff CD8+ T cells" = "#5c4f3d",
                     "NK cells" = "#b797c1",
                     "ILCs" = "#3b7527",
                     "Memory B cells" = "#ffd800",
                     "Naive B cells" = "#00abff")
```

## FlowJo Workspace

Next to some pregating steps to identify single viable cells, the manual gating
information is not used for the cluster definition of the data. However, we 
provide the manual labels for easier interpretation of the FlowSOM tree. In the 
original workspace the gates were named quite detailed, which is reduced here to
a smaller set of gate names for readibility.

```{r}
gates_of_interest <- c("Naive B cells" = "Naive B cells", 
                       "Transitional B cells" = "Naive B cells",
                       "Non Switched B cells" = "Memory B cells", 
                       "Switched Memory B cells" = "Memory B cells",
                       "CD16hi NK cells" = "NK cells", 
                       "CD56 dim NK cells" = "NK cells", 
                       "CD56hi NK cells" = "NK cells",
                       "CD14+ CD16+ Monocytes" = "Monocytes",
                       "CD14+ Monocytes" = "Monocytes", 
                       "CD16+ Monocytes" = "Monocytes",
                       "Basophils" = "Basophils", 
                       "ILCs/ILCs" = "ILCs",
                       "DC1" = "cDCs",
                       "DC2" = "cDCs", 
                       "Plasmacytoid DCs" = "pDCs", 
                       "CD4+ CD8+ T cells" = "Eff CD8+ T cells",
                       "CM CD4+ T cells" = "Eff CD4+ T cells", 
                       "EM CD4+ T cells" = "Eff CD4+ T cells",
                       "Naive CD4+ T cells" = "Na CD4+ T cells",
                       "T regs" = "Tregs",
                       "CM CD8+ T cells" = "Eff CD8+ T cells-",
                       "EM CD8+ T cells" = "Eff CD8+ T cells",
                       "Naive CD8+ T cells" = "Na CD8+ T cells",
                       "DN T cells" = "DN T cells",
                       "TCRgd+ T cells" = "gd T cells",
                       "Va24Ja18+ T cells/EM DN T cells" = "DN T cells",
                       "Va24Ja18+ T cells/Naive DN T cells" = "DN T cells",
                       "Va7-2+ T cells" = "Unknown",
                       "MAIT cells" = "MAIT cells")
```

# Preprocessing

We create a directory to save the preprocessed files

```{r}
if(!dir.exists("preprocessed")){
  dir.create("preprocessed")
}
```

For every sample, we identify which cells fall in the viable single cell
gate as manually gated in FlowJo 10 and extract the compensation and 
transformation as set up in the FlowJo workspaces. We did observe some cells 
with unrealistic high or low values when measured on the FACSymphony. 
Therefore, we also remove all cells in which any of the values fall out of the 
expected range. This final selection with compensated and transformed values 
are saved to new fcs files.  Additionally, the corresponding manual labels are 
also saved.

```{r}
for (sample in sample_ids) {
  print(paste0("Preprocessing ", sample))
  
  file <- fcs_files[sample]
  
  # Read in the FlowJo workspace for correct compensation, transformation,
  # pregating and further gating information
  wsp_file <- file.path(fcs_dir,
                        paste0("manual_pregating_", sample, ".wsp"))
  flowjo_res <- parse_flowjo(file, wsp_file)
  in_start_gate <- flowjo_res$gates[,"Viable cells"]
  
  # Extra step to remove symphony outliers, based on raw files
  ff <- read.FCS(file.path(fcs_dir, file),
                 truncate_max_range = FALSE)
  in_range <- (rowSums(ff@exprs >= 262140) == 0) &
    (rowSums(ff@exprs <= -111) == 0) &
    (rowSums(flowjo_res$flowFrame@exprs <= -100) == 0)
  
  # Combine the two selection criteria
  selection <- in_start_gate & in_range
  
  # Write selection to new file
  write.FCS(flowjo_res$flowFrame[selection],
            paste0("preprocessed/", file))
  
  # Extract manual labels for easier interpretation of results
  manual <- list(matrix = flowjo_res$gates[selection, ],
                 manual = manual_vector(flowjo_res$gates, 
                                        names(gates_of_interest))[selection])
  saveRDS(manual,
          paste0("preprocessed/",gsub(".fcs", "_manual.RDS", file)))
  
  # Save which cells were selected, to allow mapping back to the original
  # fcs files
  saveRDS(selection,
          paste0("preprocessed/",gsub(".fcs", "_selection.RDS", file)))
}
```

We create an aggregate file containing a random subsample of 300 000 cells from
each of the files. This subset will be used to train the FlowSOM algorithm.

```{r}
set.seed(1)
ff <- AggregateFlowFrames(paste0("preprocessed/", fcs_files),
                          cTotal = 300000 * length(fcs_files),
                          writeOutput = TRUE,
                          outputFile = "preprocessed/aggregate.fcs",
                          writeMeta = TRUE)
table(ff@exprs[,"File"])
```

# FlowSOM analysis

## Learn the tree

We train a 15 x 15 FlowSOM grid on the aggregated file using 25 channels.
To ensure optimal training for this gridsize and number of dimensions, we 
increased the number of iterations over the dataset to 20. On the resulting 
tree, 50 metaclusters are identified. We do not rescale the data (using the
numeric values as defined by the FlowJo transformation).

```{r}
ff <- read.FCS("preprocessed/aggregate.fcs")

markers_to_use <- c("CD3", "CD4", "CD8", "CD11b", "CD11c", "CD14", "CD16",
                    "CD20", "CD24_TCRgd", "CD25", "CD27", "CD38", "CD45RA",
                    "CD56", "CD123", "CD127", "CD183", "CD185", "CD194", 
                    "CD196", "CD197", "CD294|XCR1", "PD1", "IgD_Va7-2", 
                    "HLA-DR")

channels_to_use <- suppressWarnings(get_channels(ff, 
                                                 paste0(markers_to_use,"#")))

fsom <- FlowSOM(ff,
                scale = FALSE,
                xdim = 15,
                ydim = 15,
                colsToUse = channels_to_use,
                rlen = 20,
                nClus = 50,
                seed = 1)
fsom$FlowSOM$prettyColnames <- gsub(" \\(.*", "", fsom$FlowSOM$prettyColnames)
saveRDS(fsom, "FSOM.RDS")
```

```{r}
fsom <- readRDS("FSOM.RDS")
```

## Labeling the tree nodes

For easier interpretation of the tree, we determine the most represented manual
label for each of the metaclusters.

```{r}
# Get the manual labels for the subset of cells used to train the FlowSOM grid
manual_subset <- c()
for (file in fcs_files) {
  manual <- readRDS(file.path("preprocessed", 
                              gsub(".fcs", "_manual.RDS", 
                                   file)))
  selection <- read.table(file.path("preprocessed",
                                    gsub(".fcs", "_selected_aggregate.txt", 
                                         file)))
  labels <- as.character(manual$manual[selection$x])
  
  manual_subset <- c(manual_subset,
                     gates_of_interest[labels])
}
manual_subset[is.na(manual_subset)] <- "Unknown"
manual_subset <- factor(manual_subset,
                        levels = names(celltype_colors))

# Determine the most occuring label per cluster
c_label <- label_clusters(fsom, 
                          manual_subset)
```

## Map the original data to the tree

Once the tree is learned, we map the full preprocessed fcs file to the tree.
This allows us to have a cluster assigned to all cells from these files, rather
than just the subset selected for the aggregate file.  

We create a directory to save the results.

```{r}  
if (!dir.exists("clustered")){
  dir.create("clustered") 
}
```

We compute the percentage of cells assigned to each of the clusters for each of 
the files. Additionally, new fcs files are generated in which two dimensions are
added, containing the cluster and metacluster numbers for each cell. This allows
further inspection of the clusters in other tools (e.g. FlowJo).

```{r}
clusterCounts <- matrix(0,
                        nrow = length(fcs_files),
                        ncol = fsom$FlowSOM$map$nNodes,
                        dimnames = list(
                          sample_ids,
                          as.character(seq_len(fsom$FlowSOM$map$nNodes))))
for (sample in sample_ids){
  file <- fcs_files[sample]
  
  # Map the full file to the FlowSOM model
  fsom_tmp <- NewData(fsom$FlowSOM, file.path("preprocessed", file))
  
  # Count the number of cells assigned to each cluster
  table_tmp <- table(GetClusters(fsom_tmp))
  clusterCounts[sample, names(table_tmp)] <- table_tmp 
  
  # Generate new fcs files in which the cluster numbers are added
  ff <- read.FCS(file.path(fcs_dir, file),
                 truncate_max_range = FALSE)
  selection <- readRDS(file.path("preprocessed",
                                 gsub(".fcs", "_selection.RDS", file)))
  clusters <- rep(0, nrow(ff))
  clusters[selection] <- GetClusters(fsom_tmp)
  metaclusters <- rep(0, nrow(ff))
  metaclusters[selection] <- GetMetaclusters(fsom_tmp, fsom$metaclustering)
  
  ff <- cbind2(ff, cbind(FlowSOM_clusters = clusters, 
                         FlowSOM_metaclusters = metaclusters))
  write.FCS(ff,
            file.path("clustered", file))
}

# Convert counts to percentages
clusterPctgs <- t(apply(clusterCounts, 1, function(x){x/sum(x)}))

# Save the result
saveRDS(clusterCounts, "FlowSOM_counts.RDS")
saveRDS(clusterPctgs, "FlowSOM_percentages.RDS")
```

```{r}
clusterCounts <- readRDS("FlowSOM_counts.RDS")
clusterPctgs <- readRDS("FlowSOM_percentages.RDS")
```

## Comparison of the patient to the healthy controls

Once all cluster percentages are determined, we compute the z-score for the
patient: the patient percentage minus the mean of the healthy controls, divided 
by the standard deviation of the healthy controls. We then also rank the 
clusters based on their z-score.

```{r}
z_scores <- apply(clusterPctgs,
              2,
              function(x){
                (x["PID630"] - mean(x[controls])) / sd(x[controls])
              })

cluster_ranking <- order(abs(z_scores), decreasing = TRUE)
  
saveRDS(z_scores, "FlowSOM_z_scores.RDS")
saveRDS(cluster_ranking, "FlowSOM_cluster_ranking.RDS")
```

# Resulting figures

Using all of the previously computed results, we generated the figures for
the paper.

```{r}
ff <- read.FCS("preprocessed/aggregate.fcs")
fsom <- readRDS("FSOM.RDS")
clusterCounts <- readRDS("FlowSOM_counts.RDS")
clusterPctgs <- readRDS("FlowSOM_percentages.RDS")
z_scores <- readRDS("FlowSOM_z_scores.RDS")
cluster_ranking <- readRDS("FlowSOM_cluster_ranking.RDS")
```

## Overview of the cluster labels

```{r fig.width=10}
pdf("FlowSOM_pies_overview.pdf", width = 8, height = 11)
PlotPies(UpdateNodeSize(fsom$FlowSOM, 
                        reset = TRUE, maxNodeSize = 4),
         manual_subset,
         colorPalette = colorRampPalette(celltype_colors))
dev.off()
```

```{r fig.width=10}
PlotDiscreteVariable(UpdateNodeSize(fsom$FlowSOM, 
                                    reset = TRUE, maxNodeSize = 5),
                     variable = factor(c_label, 
                                       levels = names(celltype_colors)),
                     colors = celltype_colors)
```

## Overview figure of the z-scores

```{r}
z_score_cutoff <- 2
diff_clusters <- which(abs(z_scores) > z_score_cutoff)
```

```{r fig.width = 12, fig.height = 4}
#ord <- order(z_scores)
#ord <- seq_along(z_scores)
ord <- order(factor(c_label, levels = names(celltype_colors)))

par(mar=c(2,4,1,1))
plot(x = 1:225, 
     y = z_scores[ord],
     col = paste0(c("#225ea8","#ef6548")[(1 + (z_scores > 0))[ord]],
                  c("33", "FF")[1+(abs(z_scores) > z_score_cutoff)[ord]]),
     pch = 19,
     ylim = c(-5, 35),
     xlab = "", xaxt = "n",
     ylab = "Z-score")
abline(h = z_score_cutoff, lty = 3)
abline(h = -z_score_cutoff, lty = 3)

text(sapply(diff_clusters, function(x) which(ord == x)),
     z_scores[diff_clusters],
     diff_clusters,
     col = gsub("#FFFFFF", "#000000", celltype_colors[c_label[diff_clusters]]),
     pos = c(1,3)[(1 + (z_scores > 0))][diff_clusters],
     cex = 1)
```

## Overview of the most different clusters

```{r fig.width = 10, fig.height = 10}
node_colors <- c("--" = "white", 
                 "Underexpressed" = "#225ea8", 
                 "Overexpressed" = "#ef6548")
node_text_colors <- c("--" = "black", 
                      "Underexpressed" = "white", 
                      "Overexpressed" = "white")
node_types <- factor(c("--", "Underexpressed", "Overexpressed")[1 + 
                                                                  (abs(z_scores) > z_score_cutoff) + 
                                                                  (z_scores > z_score_cutoff)])
graph <- fsom$FlowSOM$MST$graph
igraph::V(graph)$size <- 5
igraph::V(graph)$color <- node_colors[as.character(node_types)]
igraph::V(graph)$label <- seq_len(fsom$FlowSOM$map$nNodes)
igraph::V(graph)$label.color <- node_text_colors[as.character(node_types)]
igraph::V(graph)$label.cex <- 0.5
igraph::plot.igraph(graph, 
                    layout = fsom$FlowSOM$MST$l,
                    edge.lty = 1,
                    mark.groups = as.list(1:225), 
                    mark.col = celltype_colors[c_label], 
                    mark.border = celltype_colors[c_label],
                    mark.expand = 3)

```

## Patient distribution

```{r fig.width = 10}
PlotPies(UpdateNodeSize(fsom$FlowSOM, reset = TRUE, maxNodeSize = 5),
         sample_ids[fsom$FlowSOM$data[,"File"]],
         colorPalette = colorRampPalette(sample_colors),
         backgroundValues = factor(c_label, 
                                       levels = names(celltype_colors)),
         backgroundColor = paste0(celltype_colors, "33"))
```

## Closer inspection of two clusters

```{r fig.width = 8, fig.height=8}
layout(matrix(1:4, ncol = 2))
PlotClusters2D(fsom$FlowSOM,
               "FSC-A", "FSC-H",
               51)
PlotClusters2D(fsom$FlowSOM,
               suppressWarnings(get_channels(ff, "CD20#")), 
               suppressWarnings(get_channels(ff, "CD14#")),
               51)

PlotClusters2D(fsom$FlowSOM,
               "FSC-A", "FSC-H",
               66)
PlotClusters2D(fsom$FlowSOM,
               suppressWarnings(get_channels(ff, "CD3#")), 
               suppressWarnings(get_channels(ff, "CD14#")),
               66)
```


## Overview list of the z scores

```{r}
sort(z_scores[z_scores > z_score_cutoff], decreasing = TRUE)
sort(z_scores[z_scores < -z_score_cutoff])
```

## Box plots of the differential clusters

```{r fig.width = 6, fig.height = 10}
to_plot <- data.frame()

for(cluster in 1:225){
  to_plot <- rbind(to_plot,
                   data.frame(cluster,
                              sample = rownames(clusterPctgs), #c(rep("Ctrl", 7),"R688*/R688*"),
                              value = clusterPctgs[,cluster]))
}
to_plot$cluster <- factor(to_plot$cluster,
                          levels = ord)
to_plot$cluster_names <- paste0(to_plot$cluster, 
                                " (", c_label[as.character(to_plot$cluster)],")")
to_plot$cluster_names <- factor(to_plot$cluster_names,
                                levels = paste0(ord,
                                                " (", c_label[ord],")"))

to_plot_extra <- data.frame(t(rbind(cluster = 1:225, 
                                    sapply(1:225, function(cluster){
                                      d <- to_plot$value[to_plot$cluster == cluster &
                                                           to_plot$sample %in% controls]
                                      c(mean = mean(d), sd = sd(d))}),
                                    z_score = z_scores)))

to_plot_extra$cluster <- factor(to_plot_extra$cluster,
                                levels = ord)

to_plot_extra$cluster_names <- paste0(to_plot_extra$cluster, 
                                      " (", c_label[as.character(to_plot_extra$cluster)],")")
to_plot_extra$cluster_names <- factor(to_plot_extra$cluster_names,
                                      levels = paste0(ord,
                                                      " (", c_label[ord],")"))

p <- ggplot(to_plot[to_plot$cluster %in% diff_clusters, ]) +
  geom_point(aes(x = cluster_names, y = value, col = sample)) +
  scale_color_manual(values = sample_colors) +
  geom_crossbar(aes(x = cluster_names, y = mean, ymin = mean - sd, ymax = mean + sd),
                width = 0.5, col = "grey", 
                data = to_plot_extra[to_plot_extra$cluster %in% diff_clusters, ]) +
  geom_text(aes(x = cluster_names, label = round(z_score, 2)), y = 0.03, #angle = 90,
            data = to_plot_extra[to_plot_extra$cluster %in% diff_clusters, ]) +
  coord_flip() +
  theme_minimal() +
  theme(axis.text.y = element_text(colour = gsub("#FFFFFF", "#000000", celltype_colors[c_label[names(sort(sapply(diff_clusters, function(x){which(x == ord)})))]]))) +
  ylab("% of viable PBMCs") +
  xlab("") +
  ylim(c(0, 0.035))

ggsave("Boxplots_significant_clusters.pdf",
       width = 8.3, height = 11.7,
       plot = p)

print(p)
```

```{r}
output <- clusterPctgs
colnames(output) <- paste0(colnames(output), 
                           " (", c_label[colnames(output)], ")")
openxlsx::write.xlsx(t(output[, diff_clusters]),
                     col.names = TRUE,
                     row.names = TRUE,
                     "diff_cluster_percentages.xlsx")
```

# Regarding questions of the reviewers

## CD4+ naive T cell subset not found in manual gating

```{r, fig.width = 15}
marker_heatmap <- fsom$FlowSOM$map$medianValues[, channels_to_use]
rownames(marker_heatmap) <- as.character(seq_len(nrow(marker_heatmap)))
colnames(marker_heatmap) <- get_markers(ff, colnames(marker_heatmap))
pheatmap::pheatmap(marker_heatmap[grep("Na CD4\\+ T cells", c_label), ], 
                   scale = "column")
```

```{r fig.width = 16, fig.height=20}
clusters_to_plot <- c(154, 153, 183, 155,79)
layout(matrix(seq_len(4 * length(clusters_to_plot)), 
              nrow = length(clusters_to_plot),
              byrow = TRUE))
for(cl in clusters_to_plot){
  PlotClusters2D(fsom$FlowSOM,
                 suppressWarnings(get_channels(ff, "CD3#")), 
                 suppressWarnings(get_channels(ff, "CD4#")),
                 cl,
                 main = cl)
  PlotClusters2D(fsom$FlowSOM,
                 suppressWarnings(get_channels(ff, "CD197#")), 
                 suppressWarnings(get_channels(ff, "CD45RA#")),
                 cl,
                 main = cl)
  PlotClusters2D(fsom$FlowSOM,
                 suppressWarnings(get_channels(ff, "CD27#")), 
                 suppressWarnings(get_channels(ff, "CD38#")),
                 cl,
                 main = cl)

  PlotClusters2D(fsom$FlowSOM,
                 suppressWarnings(get_channels(ff, "PD1#")), 
                 suppressWarnings(get_channels(ff, "CD127#")),
                 cl,
                 main = cl)
}
```